## Impact Of Severe Weather Events on Public Health and Economy

## Synopsis

In this report we try to understand the impact of severe weather events on public health and economy. Many severe events can 
result in fatalities, and property damage, and preventing such outcomes to the extent possible is a key concern.  To explore 
the impact, the Storm Data available from the U.S. National Oceanic and Atmospheric Administration's (NOAA) was used for this 
report.  From these data, it was determined that the top five severe event for public health in terms of fatalities and injuries 
across U.S are tornadoes,excessive heat, flash flood, heat, and lighting.  In terms of impact on the economy, floods do most 
property damage and drought has the most impact on crop damage.  Overall, Floods, hurricane/typhoons, and tronadoes are the three 
most costly weather events.

## Loading and Processing the Raw Data
### Reading in the Data
The raw data is read in from file included in the repdata-data-StormData.csv.bz2 file provided by the project 2 assignment.  The 
data is comma delimited.  The raw data is large and reading in all data is time consuming. Only the required data is read.  This is 
done by using the colClasses of read.csv.  There are 37 columns and setting the required columns to NA tells R to use the default 
method for reading the column.
```{r}
cols = rep("NULL",37)
cols[c(8,23,24,25,26,27,28)]=NA
data <- read.csv("repdata-data-StormData.csv",colClasses=cols)
```
After reading in the data we check the first few rows.
```{r}
head(data[1:5,])
```
### Data Processing 
In this step, we try to get our data ready for analyses. The two questions we are trying to answer are which types of events 
(as indicated in the EVTYPE variable) are most harmful with respect to population health? Across the United States, which types 
of events have the greatest economic consequences?  

To answer the first, we sum up the fatalities and injuries according to event type. Create a common data frame with both fatalities
and injuries.  Sort the common data first by fatalities and then by injuries.  The reason for doing so is to give fatalities more 
importance over injuries.  To get an overall population impact, we create a new column "total" which adds both fatalities and 
injuries.    
```{r} 
fatPerEvent <- aggregate(FATALITIES ~ EVTYPE,data=data,sum)
injPerEvent <- aggregate(INJURIES ~ EVTYPE,data=data,sum)
popImpact <- merge(fatPerEvent,injPerEvent,by="EVTYPE")
popImpact$total <- popImpact$FATALITIES + popImpact$INJURIES
popImpact <- popImpact[with(popImpact,
                            order(FATALITIES,INJURIES,
                            decreasing=TRUE)),]
injPerEvent <- injPerEvent[with(injPerEvent,
                                order(INJURIES,
                                      decreasing=TRUE)),]
```
Across the United States, which types of events have the greatest economic consequences?
To prepare the data for this analyses, a subset of the data is created where the 'PROPDMGEXP' and 'CROPDMGEXP' were equal to 
"K","M","B","m","k".  These values are then mapped to dollar amounts where K/k=1e3, M/m=1e6, B=1e9 and a new column is created
with these values.  The cost for each event is figured out by mutiplying the newly created column with the actual damage. 
After this step, cost is summed up for each event and sorted from highest to lowest.  To figure out the overall impact on the 
economy, the property damage and crop damage data is merged together and a new column "total" is created to represent the cost of 
both property damage and crop damage.
```{r}
# property damage
propDamage <- subset(data,PROPDMGEXP %in% c("K","M","B","m","k"))
cropDamage <- subset(data,CROPDMGEXP %in% c("K","M","B","m","k"))

mapDollar <- data.frame(exp=factor(c("K","M","B","m","k")),
                        value=c(1e3,1e6,1e9,1e6,1e3))
propDamage$dollarCost <- mapDollar$value[
                          match(propDamage$PROPDMGEXP,
                                mapDollar$exp)]

propDamage$dollarCost <- propDamage$dollarCost * propDamage$PROPDMG

propDamagePerType <- aggregate(dollarCost ~ EVTYPE,
                               data=propDamage,sum)
propDamagePerType <- propDamagePerType[with(propDamagePerType,
                                            order(dollarCost,
                                                  decreasing=TRUE)
                                            ),
                                       ]
#crop damage cost
cropDamage$dollarCost <- mapDollar$value[
  match(cropDamage$CROPDMGEXP,
        mapDollar$exp)]

cropDamage$dollarCost <- cropDamage$dollarCost * cropDamage$CROPDMG

cropDamagePerType <- aggregate(dollarCost ~ EVTYPE,
                               data=cropDamage,sum)
cropDamagePerType <- cropDamagePerType[with(cropDamagePerType,
                                            order(dollarCost,
                                                  decreasing=TRUE)
                                            
                                            ),
                                      ]
economyImpact <- merge(propDamagePerType,
                       cropDamagePerType,
                       by="EVTYPE",
                       all=TRUE)
economyImpact$total <- economyImpact$dollarCost.x + 
                        economyImpact$dollarCost.y
economyImpact <- economyImpact[with(economyImpact,
                       order(total,
                             decreasing=TRUE)
                       
                      ),
                      ]
```

## Results
### Poplulation Impact

We quickly look at the top 5 events that caused the most fatalities and injuries.
```{r}
head(popImpact[1:5,],)
```
Next, we visualize the data according to fatalities, injuries, and overall impact.  For this, ggplot2 package was used to setup the
charts and gridExtra package was used to lay them next to each other.  Here, only the top 10 weather events are plotted to get a 
feel for the most impactful events.
```{r}
require(ggplot2)
p <- ggplot(popImpact[1:10,],
            aes(x=EVTYPE,y=FATALITIES,fill=EVTYPE),height=580)
p <- p + geom_bar(stat="identity",position="dodge",colour="black")+
  guides(fill=FALSE) +
  xlab("Event Type") + 
  ylab("Number Of Fatalities") +
  ggtitle("Top 10 Events for Fatalities")  +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

p2 <- ggplot(injPerEvent[1:10,],
             aes(x=EVTYPE,y=INJURIES,fill=EVTYPE),
             height=580)
p2 <- p2 + geom_bar(stat="identity",position="dodge",colour="black")+
  guides(fill=FALSE) +
  xlab("Event Type") + 
  ylab("Number Of Injuries") +
  ggtitle("Top 10 Events for Injuries")  +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

p3 <- ggplot(popImpact[1:10,],
             aes(x=EVTYPE,y=total,fill=EVTYPE),
             height=580)
p3 <- p3 + geom_bar(stat="identity",position="dodge",colour="black")+
  guides(fill=FALSE) +
  xlab("Event Type") + 
  ylab("total impact") +
  ggtitle("Top 10 Events for overall impact")  +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
require(gridExtra)
grid.arrange(p,p2,p3,ncol=2)
```
After visualizing the data, we can clearly see that the tornadoes have the most population impact.

### Economy Impact
We quickly look at the impact of top 5 events on property damage, crop damage, and overall economy.
```{r}
head(propDamagePerType[1:5,],)
head(cropDamagePerType[1:5,],)
head(economyImpact[1:5,],)
```
Next, we visualize the data according to dollar amount per event. For this, ggplot2 package was used to setup the
charts and gridExtra package was used to lay them next to each other.  Here, only the top 10 weather events are plotted to get a 
feel for the most impactful events.

```{r}
# visualize - plot data
ep <- ggplot(propDamagePerType[1:10,],
            aes(x=EVTYPE,y=dollarCost,fill=EVTYPE),height=580)
ep <- ep + geom_bar(stat="identity",position="dodge",colour="black")+
  guides(fill=FALSE) +
  xlab("Event Type") + 
  ylab("property damage($)") +
  ggtitle("Prop Damage(Top 10)")  +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ep2 <- ggplot(cropDamagePerType[1:10,],
             aes(x=EVTYPE,y=dollarCost,fill=EVTYPE),
             height=580)
ep2 <- ep2 + geom_bar(stat="identity",position="dodge",colour="black")+
  guides(fill=FALSE) +
  xlab("Event Type") + 
  ylab("crop damage($)") +
  ggtitle("Crop Damage (Top Ten)")  +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ep3 <- ggplot(economyImpact[1:10,],
             aes(x=EVTYPE,y=total,fill=EVTYPE),
             height=580)
ep3 <- ep3 + geom_bar(stat="identity",position="dodge",colour="black")+
  guides(fill=FALSE) +
  xlab("Event Type") + 
  ylab("total impact($)") +
  ggtitle("Economy Impact (Top 10) ")  +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
require(gridExtra)
grid.arrange(ep,ep2,ep3,ncol=2)
```
From the charts we can see that the floods cause the most property damage, drought the most crop damage.  Floods, Huricane/Typhoon,
and tornadoes are the top three when it comes to impacting the overall economy.




